Visualisation of Brain Statistics with R-packages \emph{ggseg} and \emph{ggseg3d}

### 2019-12-17 msc_ggseg.tex
preprint submission

\affil[1]{Center for Lifespan Changes in Brain and Cognition, Univeristy of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway}

@@

\begin{abstract}
There is increased emphasis on visualizing results from statistical analysis in more intuitive ways. Results from imaging data inherently have a spatial component. Commonly used tools to disseminate results (i.e.~bar-charts or similar) ignore this spatial component. A tool that easily integrates the spatial aspect of the data could greatly improve dissemination of results, and also increase efficiency of workflows. The ggseg-packages make it possible to visualize pre-defined brain segmentations as both 2D polygons or 3D meshes. Presenting both the relative position and size of segments in addition to the statistical summary information eases interpretation and dissemination of results. Here we present the main data and functions in the \emph{ggseg} and \emph{ggseg3d} R-packages for brain atlas visualization. The main highlighted functions create brain segmentation plots in R, and datasets included in the packages, as well as in the \emph{ggsegExtra} companion package, fascilitate these visualizations.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, thus it is desirable that any visual representation respects this relation.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in \emph{ggplot2} (\citet{ggplot}), one can plot neuroimaging data directly in R with several tools such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The \emph{ggseg} and \emph{ggseg3d} -- pluss the complimentary \emph{ggsegExtra} -- packages include pre-compiled datasets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in \emph{ggseg} is based on polygons and \emph{ggplot2}-based grammar of graphics (\citet{ggplot}), while the 3D functionality in \emph{ggseg3d} is based on tri-surface mesh plots and \emph{plotly} (\citet{plotly}).

Both packages present compiled datasets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The \emph{ggsegExtra} package contains a collection of precompiled atlases (currently 15 additional atlases) and it is updated frequently.
A summary of all available atlases compatible with the \emph{ggseg}-packages as of December 2019 can be found in \ref{tab:atlasTab}.

@@

\emph{ggseg} is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \ref{fig:init}).
The \texttt{ggseg()}-function is a wrapper for \texttt{geom\_polygon()} from \emph{ggplot2}, and it can be built upon and combined like any \emph{ggplot}-object.
The image plot consists of a simple brain representation containing no extra information.
Hence, \emph{ggseg} plots can be easily complemented with any of the available \emph{ggplot2} features and options.
We recommend users to get familiarized with \emph{ggplot2} (\citet{ggplot}).
The package is currently only available through github, but we expect to submit to The Comprehensive R Archive Network (\citet{cran}) some time during 2020.

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/init-1.pdf}
\caption{\label{fig:init}By default \texttt{ggseg()} will plot the dkt atlas in grey shaded polygons.}
\end{figure}

In addition to the standard options for \emph{ggplot2} polygon geoms, the function also has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the \texttt{dkt}, one can stack the hemispheres, view only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view (see Figure \ref{fig:collection} for examples).
For subcortical atlases, such as the \texttt{aseg} atlas, the options are more limited but one can often choose between axial, sagittal, and coronal views.

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}\emph{ggseg} plots can be used with any \emph{ggplot2} feature such as standard scales and themes. For cortical atlases, one can supply special \emph{ggseg} options to determine hemisphere or view, and control how the plots looks like and is organised. \textbf{A:} \texttt{dkt} atlas, stacked with dark theme ; \textbf{B:} \texttt{dkt} with medial view only; \textbf{C:} \texttt{dkt} atlas with only left medial display; \textbf{D:} \texttt{dkt} atlas, stacked, with classic theme; \textbf{E:} \texttt{dkt} atlas with left hemisphere only; \textbf{F:} complete \texttt{aseg} atlas}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/fill-1.pdf}
\caption{\label{fig:fill}Supplying `area' to the fill option in \texttt{ggseg()}, will use the column `area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The \texttt{dkt} atlas default palette corresponds to the FreeSurferColorLut scheme.}
\end{figure}

Most users will use \texttt{ggseg()} to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, one should explore the structure of the atlas datasets.
The atlas dataset structure will help users understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique dataset.
All datasets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

In any atlas, the column `label' is particularly useful for combining the data of interest with the \emph{ggseg}-polygons.
The column `label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in \emph{ggseg} is in a long format - that is each region has its own row - and any data of interest needs to be in this same format.
Often datasets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged in order to work with ggseg.
See below an example of wide-to-long conversion.

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/datasupp-1.pdf}
\caption{\label{fig:datasupp}Supplying data through the `.data' option in \texttt{ggseg()} will make it possible to use columns in the supplied data to aesthetical arguments (such as `fill'). The \emph{ggseg} plot can be used with any other polygon compatible function from \emph{ggplot2} or \emph{ggplot2} extentions, for instance adding title, changing the legend name and the colour scheme with standard \emph{ggplot2} functions.}
\end{figure}

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.frame includes the column `hemi' with either `right' or `left' for this to happen.
In this case, data will be merged into the atlas both by `area' and by `hemi'.

@@

There is often the need to plot a statistic of interest in different groups (e.g.~thickness or brain - cognition relationships in young or older adults).
This may be obtained also with \texttt{ggseg()}, using \emph{ggplot2}'s \texttt{facet\_wrap} or \texttt{facet\_grid}, with three guiding rules:
\textbf{1)} as before, data needs to be in long format with a column indexing which group the row corresponds to (group data should appear in seperate rows, not in separate columns).
\textbf{2)} The data needs to be grouped using \emph{dplyr}'s \texttt{group\_by()} function \emph{before} providing the data to the \texttt{ggseg()}-function.
The \texttt{ggseg()}-function will detect grouped data, and adapt it to \texttt{facet}`s requirements.
\textbf{3)} Apply \texttt{facet\_wrap} or \texttt{facet\_grid} to the plot having used the above two rules.
An example of this can be seen in Figure \ref{fig:datasupp3}, where a mock dataset including summary statistics for two groups ('Young' and `Old') is used when facetting a \emph{ggseg}-plot.

All the concepts described above also work with the \texttt{aseg} atlas for subcortical structures, with the exception of `hemisphere' and `view' arguments that are superfluous in subcortical atlases (Figure \ref{fig:atlases}).

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/datasupp3-1.pdf}
\caption{\label{fig:datasupp3}Data including different groups need to be grouped using \texttt{group\_by()} \textbf{before} plottinh with \texttt{ggseg()} for facetting to work.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/atlases-1.pdf}
\caption{\label{fig:atlases}The \texttt{aseg} atlas, showing subcortical structures, has some distinct differences from the \texttt{dkt}. For instance, there is no option to show only a single hemisphere, as the resulting plot would be very difficult to understand. Furthermore, rather than showing lateral and medial surfaces, it shows an axial and sagittal slice.}
\end{figure}

@@

Representing brains as 2D polygons is a good solution for fast, efficient, and flexible plotting, and can be easily combined with interactive apps such as Shiny (\citet{shiny}).
Yet, brains are intrinsically 3-dimensional and it can be challenging to recognize the location of a region as a flattened image.
This problem is exacerbated in atlases that represent subcortical features as they are 3-dimensional, while cortical strucctures, such as grey matter structures or white matter tracts, can be flattened to 2-dimensions.
Hence, here we also provide the \emph{ggseg3d} package to plot, view, and print 3D-atlases in R.
\emph{ggseg3d} is based on tri-surface mesh plots using \emph{plotly} (\citet{plotly}).
The data structure is more complex than the \emph{ggplot2} polygons, and includes additional options for brain inflation, glass brains, camera locations, etc.
As \emph{ggseg3d} is based on plotly, the resulting brain atlases are interactive, which guides interpretation, and is useful for public dissemination.
We recommend users to familiarize themselves with \emph{plotly} (\citet{plotly}) when using this function.

Out-of-the-box, \texttt{ggseg3d()} plots the \texttt{dkt\_3d} atlas in `LCBC' surface, but there are two more surfaces available for cortical atlases (Figure \ref{fig:ggseg3d-1-out})
The `LCBC' surface consists on a semi-inflated white matter surface based on the \emph{fsaverage} template subject.
All \texttt{{[}...{]}\_3d} atlases include a \texttt{colour} column for default colour plotting of the regions, based on the color scheme used in the neuroimaging software.

\begin{figure}[H]
\includegraphics[width=0.3\linewidth]{png/ggseg3d_white} \includegraphics[width=0.3\linewidth]{png/ggseg3d} \includegraphics[width=0.3\linewidth]{png/ggseg3d_inflated} \caption{The three surface options provided in *ggseg3d* atlases. **From left:** the 'white' surface is the white matter surface, 'LCBC' surface is the semi-inflated white matter surface (inflated over 10 iterations), and the 'inflated' surface is a fully inflated grey matter cortical surface.}\label{fig:ggseg3d-1-out}
\end{figure}

The 3D-atlas data is stored in nested tibbles.
Each cortical atlas has datasets for three different surfaces (see Figure \ref{fig:ggseg3d-1-out}) and the two hemispheres.
Only one surface is available for subcortical atlases as inflation procedures do not yield better representations.
The `ggseg\_3d' column includes all necessary information for \texttt{ggseg3d()} to create a 3D mesh-plot, and should not be modified by the user.
The additional 3D-atlases in \emph{ggsegExtra} have the same data structure.
It is important to note that the coordinates in the plot (X, Y, Z) are \textbf{not} any type of radiological coordinate system, but arbitrary Cartesian plot coordinates.

@@

Similarly as in the 2D-atlas, the user will use \texttt{ggseg3d()} to display through a colour scale some descriptive or inferential statistic.
If the data is not already in the correct long format, or uses similar naming as the atlas, The users should inspect the atlas data for a specific surface (and hemisphere, if desired) and then \texttt{unnest(ggseg\_3d)} it to see how the atlas data is organised.

@@

Note the \texttt{mesh} column, which contains lists.
Each list corresponds to a region and contains 6 vectors required to create the mesh of the tri-surface plot.
It should also be noted that the `label', `annot' and `area' columns could provide matching values for your own data.
Similarly to the \texttt{ggseg()}-function, the `label' column should match the region names used in the original neuroimaging software while `area' and `annot' provide alternative/secondary names.
It is thus important to match your regional identifiers with those used in the atlas.
To colour the segments using a column from the data, that column name needs to be supplied to the \texttt{colour} option, and providing it to the \texttt{text} option will add another line to the \emph{plotly} hover information.

@@

\begin{figure}[H]
\includegraphics[width=0.3\linewidth]{png/ggseg3d_data} \includegraphics[width=0.3\linewidth]{png/ggseg3d_colour} \includegraphics[width=0.3\linewidth]{png/ggseg3d_nacolour} \caption{**Left:** Supplying data to `ggseg3d()` works in much the same way as for `ggseg()`. Since `ggseg3d()` is based on *plotly* rather than *ggplot2*, most aesthetic adaptations must be set in the main function. Here we set the parcellation colours with the 'colour' column and accompanying mouse-hover text with the 'text' column. **Middle:** Specifying the palette for `ggseg3d()` must be done directly in the main call to the plot. The 'palette' option takes vectors of colours either as HEX-codes or R-colour names, but not a combination of these. **Right:** The colour of the `NA` values can also be changed through the option 'na.colour', which is here set to be 'black'}\label{fig:ggseg3d-data1}
\end{figure}

\hypertarget{colours}{%
\subsubsection{Colours}\label{colours}}

You can provide custom colour palettes either in hex or R-names, as seen in Figure \ref{fig:ggseg3d-data1}.
Colours will be evenly spaced when creating the colour-scale.
A palette may also be supplied as a \emph{named numeric vector}, where the vector names are the colours that users wish to use, and the numeric values are the breakpoint for each colour (e.g.~\texttt{c("red"\ =\ 0,\ "white"\ =\ 0.5,\ "blue"\ =\ 1)}).
This way the users can control the minimum and maximum values of the colour scale, and also how the gradient is applied.
If another colour than the default gray is wanted for the \texttt{NA} regions, supply `na.colour', either as HEX colour or colour name.
This option only takes a single colour.

\hypertarget{glass-brain}{%
\subsubsection{Glass brain}\label{glass-brain}}

Subcortical atlases include cortical surfaces and other landmark structures for visualization purposes only.
One can control the opacity of the these \texttt{NA} structures, to improve visualization.
Glassbrains can be added provides a frame of reference for the subcortical structures with the function \texttt{add\_glassbrain()} (Figure \ref{fig:glassbrain1}) , which takes three extra arguments: hemisphere, colour, and opacity.

@@

\begin{figure}[H]
\includegraphics[width=0.6\linewidth]{png/ggseg3d_glass} \caption{For subcortical structure visualisation, one can add a glassbrain to the plot. This will help with understanding structure locations relative to the cortex and make the plot easier to interpret. The glassbrain is controlled by three options: opacity, hemisphere, and colour.}\label{fig:glassbrain1}
\end{figure}

@@

The \emph{ggseg} and \emph{ggseg3d} packages have two atlases each, which are 2D and 3D variations of the same main atlases: the \texttt{dkt} (\citet{dkt}) and \texttt{aseg} (\citet{aseg}) atlases, default atlases used by the commonly used imaging software Freesurfer (@\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}).
These are, however, only two among many meaningful ways of segmenting the brain into different regions.
Thus, the \emph{ggsegExtra} package is library with additional datasets for plotting with the \emph{ggseg} and \emph{ggseg3d} packages.
We hope to expand this library through a combination of our own and community effort, as the need arises for new atlases, as brain new brain atlases are continuously being released.
A small collection of the atlases currently in the \emph{ggsegExtra} package may be viewed in Figure \ref{fig:ggsegExtra}

\begin{figure}[H]
\includegraphics[width=0.8\linewidth]{png/ggseg_extra} \includegraphics[width=0.4\linewidth]{png/ggseg3d_extra1} \includegraphics[width=0.4\linewidth]{png/ggseg3d_extra2} \caption{Four example datasets from the *ggsegExtra* library, plotted with the `ggseg()` and `ggseg3d()` functions.}\label{fig:ggsegExtra}
\end{figure}

@@

We thank John Muschelli for package code comments and its adaptation to neuroconductor ({[}2018{]}), and Richard Beare for the first community created atlas, and base code comments.
We are indebted to A. M. Winkler scripts for converting neuroimaging data (Freesurfer) into .ply files, a necessary step for neuroimaging to mesh-plots conversion, and to Inge Amlien for the `LCBC' surface files and helpful contributions to the ggseg3d pipeline.
We are indebted to those users that actively contributed to the package and all the individuals that worked on the framework on which the package is sustained (e.g.~R-project, neuroimaging software).
Finally, we thank Anders M. Fjell and Kristine B. Walhovd for their encouragement and financial support.

>>>>

\affil[1]{Center for Lifespan Changes in Brain and Cognition, University of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway}

@@

\begin{abstract}
There is an increased emphasis on visualizing neuroimaging results in more intuitive ways. Common statistical tools for dissemination, such as bar charts, lack the spatial dimension that is inherent in neuroimaging data. Here we present two packages for the statistical software R, \emph{ggseg} and \emph{ggseg3d}, that integrate this spatial component. The \emph{ggseg} and \emph{ggseg3d} packages visualize pre-defined brain segmentations as both 2D polygons and 3D meshes, respectively. Both packages are integrated with other well-established R-packages, allowing great flexibility. In this tutorial, we present the main data and functions in the \emph{ggseg} and \emph{ggseg3d} packages for brain atlas visualization. The main highlighted functions are able to display brain segmentation plots in R. Further, the accompanying \emph{ggsegExtra}-package includes a wider collection of atlases, and is intended for community-based efforts to develop more compatible atlases to \emph{ggseg} and \emph{ggseg3d}. Overall, the \emph{ggseg}-packages facilitate parcellation-based visualizations in R, improve and ease the dissemination of the results, and increase the efficiency of the workflows.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, thus it is desirable that any visual representation takes this relation into account.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in \emph{ggplot2} (\citet{ggplot}), one can plot neuroimaging data directly in R with several tools such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The \emph{ggseg} and \emph{ggseg3d} -- pluss the complimentary \emph{ggsegExtra} -- packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in \emph{ggseg} is based on polygons and \emph{ggplot2}-based grammar of graphics (\citet{ggplot}), while the 3D functionality in \emph{ggseg3d} is based on tri-surface mesh plots and \emph{plotly} (\citet{plotly}).

Both packages present compiled data sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The \emph{ggsegExtra} package contains a collection of precompiled atlases (currently 15 additional atlases) and it is frequently updated.
A summary of all available atlases compatible with the \emph{ggseg}-packages as of December 2019 can be found in Table \ref{tab:atlasTab}.

@@

\emph{ggseg} is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \ref{fig:init}).
The \texttt{ggseg()}-function is a wrapper for \texttt{geom\_polygon()} from \emph{ggplot2}, and it can be built upon and combined like any \emph{ggplot}-object.
The image plot consists of a simple brain representation containing no extra information.
Hence, \emph{ggseg} plots can be easily complemented with any of the available \emph{ggplot2} features and options.
We recommend users to get familiarized with \emph{ggplot2} (\citet{ggplot}).
The package is currently only available through github, but we expect to submit the \emph{ggseg}-package to The Comprehensive R Archive Network (\citet{cran}) in 2020.

@@

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/init-1.pdf}
\caption{\label{fig:init}By default \texttt{ggseg()} will plot the dkt atlas in grey shaded polygons.}
\end{figure}

In addition to the standard options for \emph{ggplot2} polygon geoms, the function also has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the \texttt{dkt}, one can stack the hemispheres, display only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view (see Figure \ref{fig:collection} for examples).
For subcortical atlases, such as the \texttt{aseg} atlas, the options are more limited but one can often choose between axial, sagittal, and coronal views.

@@

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}\emph{ggseg} plots can be used with any \emph{ggplot2} feature such as standard scales and themes. For cortical atlases, one can supply special \emph{ggseg} options to determine, for example, hemisphere, view, or position. \textbf{A:} \texttt{dkt} atlas, stacked with dark theme ; \textbf{B:} \texttt{dkt} with medial view only; \textbf{C:} \texttt{dkt} atlas with only left medial display; \textbf{D:} \texttt{dkt} atlas, stacked, with classic theme; \textbf{E:} \texttt{dkt} atlas with left hemisphere only; \textbf{F:} complete \texttt{aseg} atlas}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/fill-1.pdf}
\caption{\label{fig:fill}Supplying `area' to the fill option in \texttt{ggseg()}, will use the column `area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The \texttt{dkt} atlas default palette corresponds to the FreeSurferColorLut scheme.}
\end{figure}

Most users will use \texttt{ggseg()} to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, one should explore the structure of the atlas data sets.
The atlas data set structure will help users understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique data set.
All data sets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Look at the top 5 rows of the dkt dataset}
\KeywordTok{head}\NormalTok{(dkt, }\DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

In any atlas, the column `label' is particularly useful for combining the data of interest with the \emph{ggseg}-polygons.
The column `label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in \emph{ggseg} is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
See below an example of wide-to-long conversion.

@@

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/datasupp-1.pdf}
\caption{\label{fig:datasupp}Supplying data through the `.data' option in \texttt{ggseg()} enables use of columns in the supplied data to aesthetical arguments (such as `fill'). The \emph{ggseg} plot can be used with any other polygon compatible function from \emph{ggplot2} or \emph{ggplot2} extentions, for instance adding title, changing the legend name and the colour scheme with standard \emph{ggplot2} functions.}
\end{figure}

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.frame includes the column `hemi' with either `right' or `left'.
In this case, data will be merged into the atlas both by `area' and by `hemi'.
For more information about adapting data and viewing only one hemisphere or side, the \href{https://lcbc-uio.github.io/ggseg/articles/ggseg.html\#single-hemisphere-results}{package vignettes} contains more elaborate information.

@@

There is often the need to plot a statistic of interest in different groups (e.g.~thickness or brain - cognition relationships in young or older adults).
This may be obtained also with \texttt{ggseg()}, using \emph{ggplot2}'s \texttt{facet\_wrap} or \texttt{facet\_grid}, with three guiding rules:
\textbf{1)} as before, data needs to be in long format with a column indexing which group the row corresponds to (group data should appear in seperate rows, not in separate columns).
\textbf{2)} The data needs to be grouped using \emph{dplyr}'s \texttt{group\_by()} function \emph{before} providing the data to the \texttt{ggseg()}-function.
The \texttt{ggseg()}-function will detect grouped data, and adapt it to \texttt{facet}`s requirements.
\textbf{3)} Apply \texttt{facet\_wrap} or \texttt{facet\_grid} to the plot having used the above two rules.
An example of this can be seen in Figure \ref{fig:datasupp3}, where a mock data set including summary statistics for two groups ('Young' and `Old') is used when faceting a \emph{ggseg}-plot.

All the concepts described above also work with the \texttt{aseg} atlas for subcortical structures, except for `hemisphere' and `view' arguments that are superfluous in subcortical atlases (Figure \ref{fig:atlases}).

@@

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/datasupp3-1.pdf}
\caption{\label{fig:datasupp3}To visualize data corresponding to different groups, one needs to use \texttt{group\_by} or similar \textbf{before} plotting with \texttt{ggseg()} for facetting to work.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/atlases-1.pdf}
\caption{\label{fig:atlases}The \texttt{aseg} atlas showing subcortical structures has some distinct differences from the \texttt{dkt}. For instance, there is no option to show only a single hemisphere. Furthermore, rather than showing lateral and medial surfaces, it shows an axial and sagittal slice.}
\end{figure}

@@

Representing brains as 2D polygons is a good solution for fast, efficient, and flexible plotting, and can be easily combined with interactive apps such as Shiny (\citet{shiny}).
Yet, brains are intrinsically 3-dimensional and it can be challenging to recognize the location of a region as a flattened image.
This problem is exacerbated in atlases that represent subcortical features as they are 3-dimensional, while cortical structures, such as grey matter structures, can be flattened to 2-dimensions.
Hence, here we also provide the \emph{ggseg3d} package to plot, view, and print 3D-atlases in R.
\emph{ggseg3d} is based on tri-surface mesh plots using \emph{plotly} (\citet{plotly}).
The data structure is more complex than the \emph{ggplot2} polygons, and includes additional options for brain inflation, glass brains, camera locations, etc.
As \emph{ggseg3d} is based on plotly, the resulting brain atlases are interactive, which guides interpretation, and is useful for public dissemination.
We recommend users to familiarize themselves with \emph{plotly} (\citet{plotly}) when using this function.

Out-of-the-box, \texttt{ggseg3d()} plots the \texttt{dkt\_3d} atlas in `LCBC' surface, but there are two more surfaces available for cortical atlases (Figure \ref{fig:ggseg3d-1-out})
The `LCBC' surface consists on a semi-inflated white matter surface based on the \emph{fsaverage5} template subject.
All \texttt{{[}...{]}\_3d} atlases include a \texttt{colour} column that based on the color scheme used in the source neuroimaging software.

\begin{figure}[H]
\includegraphics[width=0.3\linewidth]{png/ggseg3d_white} \includegraphics[width=0.3\linewidth]{png/ggseg3d} \includegraphics[width=0.3\linewidth]{png/ggseg3d_inflated} \caption{The three surface options provided in \emph{ggseg3d} atlases. \textbf{From left to right:} the `white' surface is the white matter surface, `LCBC' surface is the semi-inflated white matter surface (inflated over 10 iterations), and the `inflated' surface is a inflated grey matter cortical surface as provided by the FreeSurfer software.}\label{fig:ggseg3d-1-out}
\end{figure}

The 3D-atlas data is stored in nested tibbles.
Each cortical atlas has data sets for three different surfaces (see Figure \ref{fig:ggseg3d-1-out}) and the two hemispheres.
Only one surface is available for subcortical atlases as inflation procedures are irrelevant.
The `ggseg\_3d' column includes all necessary information for \texttt{ggseg3d()} to create a 3D mesh-plot, and should not be modified by the user.
The additional 3D-atlases in \emph{ggsegExtra} have the same data structure.
It is important to note that the coordinates in the plot (X, Y, Z) are \textbf{not} any type of radiological coordinate system, but arbitrary Cartesian plot coordinates.

@@

Similarly as in the 2D-atlas, the user will use \texttt{ggseg3d()} to display through a colour scale some descriptive or inferential statistic.
If the data is not already in the correct long format or uses similar naming as the atlas, the users should inspect the atlas data for a specific surface (and hemisphere, if desired), and then \texttt{unnest(ggseg\_3d)} it to see how the atlas data is organised.

@@

Note the \texttt{mesh} column, which contains lists.
Each list corresponds to a region and contains 6 vectors required to create the mesh of the tri-surface plot.
It should also be noted that the `label', `annot' and `area' columns could provide matching values for your own data.
Similarly to the \texttt{ggseg()}-function, the `label' column should match the region names used in the original neuroimaging software while `area' and `annot' provide alternative/secondary names.
It is thus important to match your regional identifiers with those used in the atlas.
To colour the segments using a column from the data, a column name from the data needs to be supplied to the \texttt{colour} option, and providing it to the \texttt{text} option will add another line to the \emph{plotly} hover information.

@@

\begin{figure}[H]
\includegraphics[width=0.3\linewidth]{png/ggseg3d_data} \includegraphics[width=0.3\linewidth]{png/ggseg3d_colour} \includegraphics[width=0.3\linewidth]{png/ggseg3d_nacolour} \caption{\textbf{Left:} Supplying data to \texttt{ggseg3d()} works similarly to \texttt{ggseg()}. Since \texttt{ggseg3d()} is based on \emph{plotly} rather than \emph{ggplot2}, most aesthetic adaptations must be set in the main function. Here we set the parcellation colours with the `colour' column and accompanying mouse-hover text with the `text' column. \textbf{Middle:} The palette for \emph{ggseg3d} needs to be specified directly in the main plot call. The `palette' option takes vectors of colours either as HEX-codes or R-colour names. \textbf{Right:} The colour of the \texttt{NA} values can also be changed through the option `na.colour'.}\label{fig:ggseg3d-data1}
\end{figure}

\hypertarget{customizing-colours-and-the-colour-bar}{%
\subsubsection{Customizing colours and the colour bar}\label{customizing-colours-and-the-colour-bar}}

You can provide custom colour palettes either in hex or R-names, as seen in Figure \ref{fig:ggseg3d-data1}.
Colours will be evenly spaced when creating the colour-scale.
A palette may also be supplied as a \emph{named numeric vector}, where the vector names are the colours that users wish to use, and the numeric values are the breakpoints for each colour (e.g.~\texttt{c("red"\ =\ 0,\ "white"\ =\ 0.5,\ "blue"\ =\ 1)}).
This way the users can control the minimum and maximum values of the colour scale, and also how the gradient is applied.
If another colour than the default gray is wanted for the \texttt{NA} regions, supply `na.colour', either as HEX colour or colour name.
This option only takes a single colour.

\hypertarget{adding-a-glass-brain}{%
\subsubsection{Adding a glass brain}\label{adding-a-glass-brain}}

Subcortical atlases include cortical surfaces and other landmark structures for visualization purposes only.
One can control the opacity of the these \texttt{NA} structures, to improve visualization.
Glass brains can be added to provide a frame of reference for the subcortical structures with the function \texttt{add\_glassbrain()} (Figure \ref{fig:glassbrain1}) , which takes three extra arguments: hemisphere, colour, and opacity.

@@

\begin{figure}[H]
\includegraphics[width=0.6\linewidth]{png/ggseg3d_glass} \caption{For subcortical structure visualization, one can add a glass brain to the plot. This will help with locating the structures relative to the cortex, and make the plot easier to interpret. The glass brain is controlled by three options: opacity, hemisphere, and colour.}\label{fig:glassbrain1}
\end{figure}

@@

The \emph{ggseg} and \emph{ggseg3d} packages have two atlases each, which are 2D and 3D variations of the same main atlases: the \texttt{dkt} (\citet{dkt}) and \texttt{aseg} (\citet{aseg}) atlases.
These are, however, only two among many meaningful ways of segmenting the brain into different regions.
Thus, the \emph{ggsegExtra} package is a repository containing additional with additional data sets for plotting with the \emph{ggseg} and \emph{ggseg3d} packages.
There is an ever-increasing amount of new atlases being created, as research and methods in neuroimaging analysis progresses.
The \emph{ggsegExtra}-package is intended to be expanded as a community-effort, as new and informative atlases are published.
A small collection of the atlases currently in the \emph{ggsegExtra} package may be viewed in Figure \ref{fig:ggsegExtra}

\begin{figure}[H]
\includegraphics[width=0.8\linewidth]{png/ggseg_extra} \includegraphics[width=0.4\linewidth]{png/ggseg3d_extra1} \includegraphics[width=0.4\linewidth]{png/ggseg3d_extra2} \caption{Four example data sets from the \emph{ggsegExtra} library, plotted with the \texttt{ggseg()} and \texttt{ggseg3d()} functions.}\label{fig:ggsegExtra}
\end{figure}

@@

We thank John Muschelli for package code comments and its adaptation to neuroconductor ({[}2018{]}), and Richard Beare for the first community created atlas, and base code comments.
We are indebted to A. M. Winkler scripts for converting neuroimaging data (Freesurfer) into .ply files, a necessary step for neuroimaging to mesh-plots conversion, and to Inge Amlien for the `LCBC' surface files and helpful contributions to the ggseg3d pipeline.
We are indebted to those users that actively contributed to the package and all the individuals that worked on the framework on which the package is sustained (e.g.~R-project, neuroimaging software).
Finally, we thank Anders M. Fjell (AMF) and Kristine B. Walhovd (KBW) for their encouragement and financial support.
